A comprehensive framework to capture the arcana of neuroimaging analysis

### 2019-03-18 manuscript/manuscript.tex
changed reference to specification tables to just specifications

The set of input data, the derivatives that can be derived from it,
and methods that construct pipelines to derive the derivatives, are
linked together by the \emph{data specification} \emph{table} class
attribute of the Study (Figure \ref{fig:example_study}). Likewise,
free parameters used in pipeline construction are defined in the class'
\emph{parameter specification} \emph{table}. Class inheritance can
be used to specialise analysis suites by overriding entries in the
specification tables or pipeline constructor methods. Analysis suites
for multi-modal data can be implemented by combining Study classes
within \emph{MultiStudy} classes\emph{.}

@@

Analysis methods defined by a Study class are applied to a
specific dataset by instantiating an object of the class and
requesting a derivative listed in the class' data specification table.
At initialisation, a Study object is passed references to a \emph{Repository},
a \emph{Processor}, and an \emph{Environment} objects, which
define where and how data is stored and processed. When a derivative is
requested, a Study object queries the Repository for intermediate
derivatives that can be reused before constructing a workflow to produce
the requested derivative. The manner in which the workflows are executed
(i.e. single/multi-process or via SLURM scheduler) is specified by the
Processor and software modules required by the analysis are
loaded by the Environment. Selected workflow products are
stored back in the repository for reuse by subsequent analyses (Figure
\ref{fig:simple_uml}). Input data to a study are selected from repositories
 using criteria defined in \emph{Input} objects passed to the Study
 object at initialisation and matched against entries in the class' data
specification table.

@@

\subsubsection*{Data and parameter specification
tables}
\label{sec:data-and-parameter-specification-tables}

At the heart of each Study class is the data specification table, which
is stored in the \emph{\_data\_specs} class attribute and specifies the
input and output data of the analysis, and all
stages in between. There is a one-to-one relationship between entries in
the data specification table and data that are stored in the repository,
or will be stored if and when they are derived. Which intermediate
derivatives to include in the data specification table, and therefore
store in the repository, is left to the discretion of the researcher
designing the analysis. However, as a general rule, derivatives that
require manual QC or are likely to be reused between different branches
of analysis should be included in the table.

Data specified in the data specification table can be of either a
\emph{fileset} or \emph{field} type. Filesets represent single files,
or sets of related files typically contained within a directory
(e.g. a multi-volume DICOM dataset). Fields
represent integer, floating point, datetime, or character string
values. By default, a Field represents a single value but if the \emph{array}
flag is set, a Field represents a list of values.
Each Fileset references a \emph{FileFormat} object, which
specifies the formats of the files in the set. File formats are
explicitly registered by the researcher at design time using the
\emph{FileFormat.register(format)} class method to avoid conflicts where
the same extension is used for different formats in different contexts.

Fileset and field specifications of a Study class are defined by the
list of \emph{FilesetSpec} and \emph{FieldSpec} objects in the
\emph{add\_data\_specs} class attribute (Figure \ref{fig:data_spec}),
The specifications are collated with any data specifications in base
classes into the \emph{\_data\_specs} class attribute during the creation
of the class by the StudyMetaClass metaclass.
Specifications for input data to the study are
distinguished from derivative specifications by the \emph{FilesetInputSpec}
and \emph{FieldInputSpec} subtypes. However, the distinction between
input and derived data is somewhat blurred, as derivative specifications can be
passed input data when the class is instantiated. In addition,
when subclassing a Study class, or combining it in a MultiStudy class,
derivative specifications can be overridden by input
specifications, and vice-versa.

@@

For Study classes that correspond to a known type in the BIDS
standard, a dictionary mapping their data specifications to default
\emph{BidsInput} or \emph{BidsAssocInput} selectors can be specified in the
\emph{default\_bids\_inputs} class attribute. BidsInput
specifies a primary scan in the BIDS standard using its \emph{type}, \emph{modality},
\emph{format} and optionally the \emph{task} it belongs to. BidsAssocInput is used to select
associated files, such as field maps and diffusion encoding matrices relative
to a primary scan. Default BIDS selectors are typically not provided with
a value for the \emph{task} keyword, which can be set at instantiation of
the Study class with the \emph{bids\_task} keyword argument to enable the design of Study
classes that are applicable to any BIDS task.

@@

Workflows are implemented in Arcana as a series of modular pipelines, which
each perform a unit of the analysis (e.g. registration, brain
extraction, quantitative susceptibility mapping). Pipelines are represented by \emph{Pipeline} objects,
which are thin wrappers around Nipype workflows to handle input and output
connections, and namespace management. Each pipeline consists of a
(typically small) graph of Nipype nodes, with each node wrapping a stand alone tool
(e.g. FSL's FLIRT) or analysis package function (e.g. SPM's \emph{coreg}
tool). Pipelines source their inputs from, and sink their outputs to, entries
in the data specification table, thereby linking the table together.

@@

The syntax for pipeline construction is inspired by changes proposed for Nipype v2.0
(\url{github.com/nipy/nipype/issues/2539}).
Within a pipeline constructor method, a Pipeline object is constructed by the \emph{Study.pipeline} method
(Figure \ref{fig:pipeline_constructor}). At initialisation, each pipeline is given a name,
which must be unique amongst the pipelines constructed by the Study. The methods implemented
by the pipeline can be characterised by providing a the list of citations
and a text description to the \emph{citations} and \emph{desc} keyword arguments, respectively.

@@

Node interfaces are connected to each other, and to inputs and outputs of the pipeline, by
providing \emph{inputs} and \emph{outputs} keyword arguments to the \emph{add} method. Both
arguments take a dictionary. The keys of the \emph{inputs} dictionary correspond to
trait names in the node's input interface, whereas the keys of the \emph{outputs} dictionary correspond to
names of entries in the data specification table. The values of both dictionaries are 2-tuples. For pipeline
inputs, values of the \emph{inputs} dictionary consist of a name of an entry in the data specification table
and the format the input data is expected in (i.e. a FileFormat for Fileset specifications or Python datatype for Field specifications).
For pipeline outputs, values of the \emph{outputs} dictionary consist of a trait name in the node's output interface and the format the
output data is generated in. For input connections from other nodes, values of the \emph{inputs} dictionary
consist of a reference to the upstream node and the name of a trait in the
upstream node's output interface (output connections to other nodes are implied by input connections of the receiving nodes).

@@

Instead of being set directly, the data and parameter specifications are set
by the metaclass of the Study (i.e. StudyMetaClass)
in order to combine them with corresponding specifications in the class' bases. The combined
data and parameter specifications are constructed by visiting the class'
bases in reverse method resolution order (MRO) and adding specifications
from their \emph{add\_data\_specs}, \emph{add\_param\_specs} attributes,
overriding previously added specifications with matching names. Note
that in this scheme, specifications can only be appended or overridden
but not removed by Study subclasses so as not to break workflows
inherited from base classes.

Pipeline constructor methods can be overridden in subclasses like any Python
method. Often the overriding method will call the superclass method
to construct a pipeline, apply modifications and return the
modified pipeline. In this scenario, references to the data
specification in the superclass method can be mapped onto different
entries in the table of the subclass by providing the \emph{input\_map} and
\emph{output\_map} keyword arguments when calling the superclass method.
Additionally the name of the pipeline can be altered by providing the \emph{name}
argument, which is useful when creating multiple pipeline constructor methods
from a single base method. In these scenarios it is important to
also pass the the wildcard keyword arguments of the overriding method to the
\emph{name\_maps} keyword argument of the superclass method to allow the
overriding method to be overridden in turn.

@@

While basic Study classes are typically associated with single image
modality or contrast, the analysis suites implemented by them can be
integrated into multi-modal analysis by aggregating multiple Study
classes (sub-studies) in a \emph{MultiStudy} class. Analysis suites are
integrated in a MultiStudy class by linking the data specification tables
of sub-studies. Sub-study specification table entries are linked by mapping
them to a common entry in the specification table class attribute
of the MultiStudy (Figure \ref{fig:example_multi_study}). This enables derivatives from one
sub-study (e.g. brain extracted T1-weighted anatomical) to be referenced
by workflows of other sub-studies (e.g. anatomically constrained DWI
tractography).

@@

Sub-studies are aggregated in the \emph{sub-study} \emph{specification
table} of a MultiStudy class via a list of \emph{SubStudySpec} objects
in the \emph{add\_sub\_study\_specs} class attribute in the manner of
data and parameter specifications. A SubStudySpec consists of a name, a
Study class, and a \emph{name map} dictionary\emph{.} The name map
dictionary maps data and parameter specification names from the
sub-study namespace to the namespace of the MultiStudy class, i.e.
the dictionary keys refer to entries in the sub-study specification table and
the dictionary values refer to entries in the multi-study specification table.

Entries in the specification tables of sub-study classes that are not
referenced in the sub study's name map are implicitly mapped to the
MultiStudy namespace by the MultiStudyMetaClass during construction of the
MultiStudy class using the name of the sub-study as a prefix.
If the implicitly mapped specification is derived, then its associated pipeline
constructor method is also mapped into the MultiStudy namespace with a prefix.
For example, \emph{derived1} in \emph{sub\_study2} would be mapped to
\emph{sub\_study2\_derived1} along with the method
\emph{sub\_study2\_pipeline1}.

@@

Each Study instance is assigned a name, which is used to differentiate its
results from alternative analyses on the same dataset (e.g. with different
parameterisations). Parameters are set on initialisation of the Study object
along with the range of subject and visit IDs to be included in the analysis
(if they are not provided then all IDs found in the repository are included).
The remaining arguments passed to the Study object initialisation are the
Repository, Processor and Environment modules to use and a list of
\emph{Input} objects to select input data from the repository and match
it with the Study's data specification table.
(Figure \ref{fig:study_application}).

@@

While derivatives generated by a Study object are named in accordance
with the data specification of the Study class, arbitrary naming
conventions can be used for input datasets and fields. A selection
stage matches input data to entries in the data specification table of
the Study class. The criteria for these selections are passed to the
the \emph{inputs} argument of the Study object at instantiation in a
dictionary mapping data specification names to \emph{FilesetInput} and
\emph{FieldInput} objects, and are required
to match exactly one fileset or field in every session included in the study.

@@

Specific files and fields that are not stored within a repository can be
passed as inputs in \emph{FilesetCollection} and \emph{FieldCollection}
objects. Collection objects reference an entry in the data specification
table and contain a single Fileset or Field for every session (or
subject/visit/study depending on the frequency of the corresponding data
specifcation). Collection objects can be used to pass reference atlases and
templates as inputs to analyses. They can also be set as the default
for a data specification via the \emph{default} keyword
argument. However, for the sake of portability, default inputs should be
restricted to data in publically accessible repositories or those included in
standard software packages (e.g. FSL).

>>>>

The set of input data, the derivatives that can be derived from it,
and methods that construct pipelines to derive the derivatives, are
linked together by the \emph{data specification} class attribute of the
Study (Figure \ref{fig:example_study}). Likewise,
free parameters used in pipeline construction are defined in the class'
\emph{parameter specification} class attribute. Class inheritance can
be used to specialise analysis suites by overriding entries in the
specifications or pipeline constructor methods. Analysis suites
for multi-modal data can be implemented by combining Study classes
within \emph{MultiStudy} classes\emph{.}

@@

Analysis methods defined by a Study class are applied to a
specific dataset by instantiating an object of the class and
requesting a derivative listed in the class' data specification.
At initialisation, a Study object is passed references to a \emph{Repository},
a \emph{Processor}, and an \emph{Environment} objects, which
define where and how data is stored and processed. When a derivative is
requested, a Study object queries the Repository for intermediate
derivatives that can be reused before constructing a workflow to produce
the requested derivative. The manner in which the workflows are executed
(i.e. single/multi-process or via a SLURM scheduler \citep{yoo_slurm:_2003})
is specified by the Processor and software modules required by the analysis are
loaded by the Environment. Selected workflow products are
stored back in the repository for reuse by subsequent analyses (Figure
\ref{fig:simple_uml}). Input data to a study are selected from repositories
 using criteria defined in \emph{Input} objects passed to the Study
 object at initialisation and matched against entries in the class' data
specification.

@@

\subsubsection*{Data and parameter specifications}
\label{sec:data-and-parameter-specifications}

At the heart of each Study class is the data specification attribute, which
specifies the input and output data of the analysis, and all
stages in between. There is a one-to-one relationship between entries in
the data specification and derivatives that are stored in the repository,
or will be stored if and when they are derived. Which intermediate
derivatives to include in the data specification, and therefore
store in the repository, is left to the discretion of the researcher
designing the analysis. However, as a general rule, derivatives that
require manual QC or are likely to be reused between different branches
of analysis should be included in the data specification.

Entries in the data specification can be of either a
\emph{file set} or \emph{field} type. File sets represent single files,
or sets of related files typically contained within a directory
(e.g. a multi-volume DICOM dataset). Fields
represent integer, floating point, date-time, or character string
values. By default, a field represents a single value but if the \emph{array}
flag is set, a field represents a list of values.
Each file set references a \emph{FileFormat} object, which
specifies the formats of the files in the set. File formats are
explicitly registered by the researcher at design time using the
\emph{FileFormat.register(format)} class method to avoid conflicts where
the same extension is used for different formats in different contexts.

The data specification of a Study class are defined by providing a
list of \emph{FilesetSpec} and \emph{FieldSpec} objects in the
\emph{add\_data\_specs} class attribute (Figure \ref{fig:data_spec}).
The lists are collated with data specifications in any base Study
classes during the creation of the class. Input data to the study are
distinguished from derivatives by using the \emph{FilesetInputSpec}
and \emph{FieldInputSpec} subtypes. However, the distinction between
input and derived data is somewhat blurred, as derivative specifications can be
passed input data when the class is instantiated. In addition,
when subclassing a Study class, or combining it in a MultiStudy class,
derivative specifications can be overridden by input
specifications, and vice-versa.

@@

For Study classes that correspond to a known type in the BIDS
standard, a dictionary mapping their data specifications to default
\emph{BidsInput} or \emph{BidsAssocInput} selectors can be specified in the
\emph{default\_bids\_inputs} class attribute. BidsInput
specifies a primary scan in the BIDS standard using its \emph{type}, \emph{modality},
\emph{format} and optionally the \emph{task} it belongs to. BidsAssocInput is used to select
associated files, such as field maps and diffusion encoding matrices relative
to a primary scan. Default BIDS selectors are typically not provided with
a value for the \emph{task} keyword since the task can be set at instantiation of
the Study class with the \emph{bids\_task} keyword argument, which enables the design of Study
classes that are not restricted to a single task by default.

@@

Workflows are implemented in Arcana as a series of modular pipelines, which
each perform a unit of the analysis (e.g. registration, brain
extraction, quantitative susceptibility mapping). Pipelines are represented by \emph{Pipeline} objects,
which are thin wrappers around Nipype workflows to handle input and output
connections, and namespace management. Each pipeline consists of a
(typically small) graph of Nipype nodes, with each node wrapping a stand alone tool
(e.g. FSL's FLIRT) or analysis package function (e.g. SPM's \emph{coreg}
tool). Pipelines source their inputs from, and sink their outputs to, entries
in the data specification, thereby linking the specification together.

@@

The syntax for pipeline construction is inspired by changes proposed for Nipype v2.0
(\url{github.com/nipy/nipype/issues/2539}).
Within a pipeline constructor method, a Pipeline object is constructed by the \emph{Study.pipeline} method
(Figure \ref{fig:pipeline_constructor}). At initialisation, each pipeline is assigned a name,
which must be unique amongst the pipelines constructed by the Study given the any combination of
switches. The methods implemented by the pipeline can be characterised by providing a the list of citations
and a text description to the \emph{citations} and \emph{desc} keyword arguments, respectively.

@@

Node interfaces are connected to each other, and to inputs and outputs of the pipeline, by
providing \emph{inputs} and \emph{outputs} keyword arguments to the \emph{add} method. Both
arguments take a dictionary. The keys of the \emph{inputs} dictionary correspond to
trait names in the node's input interface, whereas the keys of the \emph{outputs} dictionary correspond to
names of entries in the data specification. The values of both dictionaries are 2-tuples. For pipeline
inputs, values of the \emph{inputs} dictionary consist of a name of an entry in the data specification
and the format the input data is expected in (i.e. a FileFormat for Fileset specifications or core Python type for Field specifications).
For pipeline outputs, values of the \emph{outputs} dictionary consist of a trait name in the node's output interface and the format the
output data is generated in. For input connections from other nodes, values of the \emph{inputs} dictionary
consist of a reference to the upstream node and the name of a trait in the
upstream node's output interface (output connections to other nodes are implied by input connections to the receiving nodes).

@@

Instead of being set directly, the data and parameter specifications are set
by the metaclass of the Study (i.e. StudyMetaClass)
in order to combine them with corresponding specifications in base classes.
The combined data and parameter specifications are constructed by visiting the class'
bases in reverse method resolution order (MRO) and adding specifications
from their \emph{add\_data\_specs}, \emph{add\_param\_specs} attributes,
overriding previously added specifications with matching names. Note
that in this scheme, specifications can only be appended or overridden
but not removed by Study subclasses so as not to break workflows
inherited from base classes.

Pipeline constructor methods can be overridden in subclasses like any Python
method. Often the overriding method will call the superclass method
to construct a pipeline, apply modifications and return the
modified pipeline. In this scenario, references to the data
specification in the superclass method can be mapped onto different
entries in the specification of the subclass by providing the \emph{input\_map} and
\emph{output\_map} keyword arguments when calling the superclass method.
Additionally the name of the pipeline can be altered by providing the \emph{name}
argument, which is useful when creating multiple pipeline constructor methods
from a single base method. In these scenarios it is important to
also pass the the wildcard keyword arguments of the overriding method to the
\emph{name\_maps} keyword argument of the superclass method to allow the
overriding method to be overridden in turn.

@@

While basic Study classes are typically associated with single image
modality or contrast, the analysis suites implemented by them can be
integrated into multi-modal analysis by aggregating multiple Study
classes (sub-studies) in a \emph{MultiStudy} class. Analysis suites are
integrated in a MultiStudy class by linking the data specifications
of sub-studies. Sub-study specification entries are linked by mapping
them to a common entry in the specification of the MultiStudy class
(Figure \ref{fig:example_multi_study}). This enables derivatives from one
sub-study (e.g. brain extracted T1-weighted anatomical) to be referenced
by workflows of other sub-studies (e.g. anatomically constrained DWI
tractography).

@@

Sub-studies are aggregated in the \emph{sub-study} \emph{specification}
of a MultiStudy class via a list of \emph{SubStudySpec} objects
in the \emph{add\_sub\_study\_specs} class attribute in the manner of
data and parameter specifications. A SubStudySpec consists of a name, a
Study class, and a \emph{name map} dictionary\emph{.} The name map
dictionary maps data and parameter specification names from the
sub-study namespace to the namespace of the MultiStudy class, i.e.
the dictionary keys refer to entries in the sub-study specification and
the dictionary values refer to entries in the multi-study specification.

Entries in the specifications of sub-study classes that are not
referenced in the sub study's name map are implicitly mapped to the
MultiStudy namespace by the MultiStudyMetaClass during construction of the
MultiStudy class using the name of the sub-study as a prefix.
If the implicitly mapped specification is derived, then its associated pipeline
constructor method is also mapped into the MultiStudy namespace with a prefix.
For example, \emph{derived1} in \emph{sub\_study2} would be mapped to
\emph{sub\_study2\_derived1} along with the method
\emph{sub\_study2\_pipeline1}.

@@

Each Study instance is assigned a name, which is used to differentiate its
results from alternative analyses on the same dataset (e.g. with different
parameterisations). Parameters are set on initialisation of the Study object
along with the range of subject and visit IDs to be included in the analysis
(if they are not provided then all IDs found in the repository are included).
The remaining arguments passed to the Study object initialisation are the
Repository, Processor and Environment modules to use and a list of
\emph{Input} objects to select input data from the repository and match
it with the Study's data specification.
(Figure \ref{fig:study_application}).

@@

While derivatives generated by a Study object are named in accordance
with the data specification of the Study class, arbitrary naming
conventions can be used for input datasets and fields. A selection
stage matches input data to entries in the data specification of
the Study class. The criteria for these selections are passed to the
the \emph{inputs} argument of the Study object at instantiation in a
dictionary mapping data specification names to \emph{FilesetInput} and
\emph{FieldInput} objects, and are required
to match exactly one fileset or field in every session included in the study.

@@

Specific files and fields that are not stored within a repository can be
passed as inputs in \emph{FilesetCollection} and \emph{FieldCollection}
objects. Collection objects reference an entry in the data specification
and contain a single Fileset or Field for every session (or
subject/visit/study depending on the frequency of the corresponding data
specifcation). Collection objects can be used to pass reference atlases and
templates as inputs to analyses. They can also be set as the default
for a data specification via the \emph{default} keyword
argument. However, for the sake of portability, default inputs should be
restricted to data in publically accessible repositories or those included in
standard software packages (e.g. FSL).
